---
title: "Example 1. Combining two independent ratings."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example 1. Combining two independent ratings.}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
references:
  - id: datasrc
    title: "Many analysts, one dataset: Making transparent how variations in anlytical choices affect results"
    type: entry
    issued:
      year: 2014
      month: 4
      day: 24
    accessed:
      year: 2019
    URL: https://osf.io/gvm2z/
    author:
    - given: Raphael 
      family: Silberzahn 
    - given: "Eric Luis" 
      family: Uhlmann 
    - given: Dan 
      family: Martin 
    - given: Pasquale 
      family: Anselmi 
    - given: Frederik 
      family: Aust 
    - given: "Eli C."
      family: Awtrey
    - given: Štěpán
      family: Bahník 
    - given: Feng
      family: Bai 
    - given: Colin
      family: Bannard
    - given: Evelina
      family: Bonnier
    - given: Rickard
      family: Carlsson
    - given: Felix
      family: Cheung
    - given: Garret
      family: Christensen
    - given: Russ
      family: Clay
    - given: "Maureen A."
      family: Craig
    - given: Anna 
      family: "Dalla Rosa"
    - given: Lammertjan
      family: Dam
    - given: "Mathew H."
      family: Evans
    - given: "Ismael Flores"
      family: Cervantes
    - given: Nathan
      family: Fong
    - given: Monica
      family: Gamez-Djokic
    - given: Andreas
      family: Glenz
    - given: Shauna
      family: Gordon-McKeon
    - given: Tim
      family: Heaton
    - given: "Karin Hederos" 
      family: Eriksson 
    - given: Moritz
      family: Heene
    - given: "Alicia Hofelich"
      family: Mohr 
    - given: Kent
      family: Hui
    - given: Magnus
      family: Johannesson
    - given: Jonathan
      family: Kalodimos
    - given: Erikson
      family: Kaszubowski
    - given: Deanna
      family: Kennedy
    - given: Ryan
      family: Lei
    - given: "Thomas Andrew"
      family: Lindsay
    - given: Silvia
      family: Liverani
    - given: Christopher
      family: Madan
    - given: "Daniel C."
      family: Molden 
    - given: Eric 
      family: Molleman
    - given: "Richard D."
      family: Morey
    - given: Laetitia
      family: Mulder
    - given: "Bernard A." 
      family: Nijstad
    - given: Bryson
      family: Pope
    - given: Nolan
      family: Pope
    - given: "Jason M."
      family: Prenoveau
    - given: Floor
      family: Rink
    - given: Egidio
      family: Robusto
    - given: Hadiya
      family: Roderique
    - given: Anna
      family: Sandberg
    - given: Elmar
      family: Schlueter
    - given: Felix
      family: S
    - given: "Martin F." 
      family: Sherman
    - given: "S. Amy"
      family: Sommer
    - given: "Kristin Lee"
      family: Sotak
    - given: "Seth M."
      family: Spain
    - given: Christoph
      family: Spörlein 
    - given: Tom
      family: Stafford
    - given: Luca
      family: Stefanutti
    - given: Susanne
      family: Täuber
    - given: Johannes
      family: Ullrich
    - given: Michelangelo 
      family: Vianello 
    - given: Eric-Jan 
      family: Wagenmakers 
    - given: Maciej 
      family: Witkowiak 
    - given: Sangsuk 
      family: Yoon 
    - given: Brian A. 
      family: Nosek
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


The library comes preloaded with
a dataset, `soccer`, from [@datasrc] 
on card counts per each pair between 
2,053 soccer players and 3,147 referees.
Below are the first few lines of the dataset.

```{r load}
library(mverse)
knitr::kable(head(soccer)) # using kable for displaying data in html
```


The dataset was used to study whether
racial bias played a role when referees
issued the cards. For each player,
two independent raters coded their skin tones
on a 5-point scale ranging 
from _very light skin_ (`0.0`)
to _very dark skin_ (`1.0`) 
with _neither dark nor light skin_(`0.5`)
as the centre value. 


We would like to study the distribution
of the player skin tones but the two independent 
rating do not always match. 
To combine the two ratings, you may consider
the following options:

1.  the mean numeric value
2.  the darker rating of the two
3.  the lighter rating of the two
4.  the first rating only
5.  the second rating only

## 1. Without the `mverse` library.

Let's first consider how you might 
study the five options using base R.  
First, we will define the five options
as separate variables.

```{r base_r}
# remove rows with no rating
soccer <- soccer[!is.na(soccer$rater1), ] 

skin_option_1 <- (                # mean
  soccer$rater1 + soccer$rater2) / 2 
skin_option_2 <- ifelse(          # max
  soccer$rater1 > soccer$rater2, 
  soccer$rater1, soccer$rater2)
skin_option_3 <- ifelse(          # min
  soccer$rater1 < soccer$rater2, 
  soccer$rater1, soccer$rater2)
skin_option_4 <- soccer$rater1    # first
skin_option_5 <- soccer$rater2    # second
```

We can plot their histograms to study the distributions.
We will use the `ggplot2` library for plotting.

```{r hist_base, fig.caption="Histogram of skin tones variale based on the first definition.", fig.width=5, fig.height=4, warning=FALSE}
library(ggplot2)
ggplot(mapping=aes(x = skin_option_1)) +
  geom_histogram(breaks = seq(0,1,0.2),
                 colour = 'white') +
  labs(title = 'Histogram of player skin tones.',
       x = 'Skin Tone', y = 'Count')
```


Now, for the remaining four variable definitions,
we will have to repeat the step above. 

### Exercise.

1.  Plot histograms for the remaining four options in base R.
How would you display the five plots together?
2.  How would you perform the same analysis 
using the `tidyverse` package?


## 2. Basic Branching using `mverse`.

We will now learn how we can
create the options using the `mverse` package.
We first consider the first two options.

1.  the mean numeric value
2.  the darker rating of the two

First, we will define a `mverse` object
from the data set.

```{r create_mv}
mv <- create_multiverse(soccer)
```

In `mverse` package, an alternative variabe definition
is called a variable branch. You can create
a variable branch with the function `variable_branch`.

```{r variable_branch}
skin_tone <- variable_branch(
  (rater1 + rater2)/2,
  ifelse(rater1 > rater2, rater1, rater2)
)
```

You can then add the variable branch rules to the `mverse` object
using `add_variable_branch`. You can use the `tidyverse`'s
`dplyr` grammar for defining the variable rules.

```{r add_vb}
mv <- mv %>%
  add_variable_branch(skin_tone) %>%
  execute_multiverse()
```

The object name `skin_tone` is the name of
the new variable defined. Each alternative
definition creates an alternative _universe_
in the `mverse` object. `execute_multiverse`
generates the new variable based
on the branching rules provided by
`add_variable_branch`.

You can see the variable branch added with `summary`
method for the `mverse` object. 

```{r check_multiverse}
summary(mv) 
```

### Exercise.

1.  Define and add the variable branches for the remaining three options.
2.  Print the multiverse table. How many _universes_ does your multiverse produce?

## 3. Comparing between Universes.

In this section, we will now examine and compare the distributions of
the variable branches. You can extract the resulting value of branched
variables using `extract` method. By default, it returns the branched
variables across all universes with the universe index. Note that
the universe index column is returned as a factor variable.

```{r extract_multiverse}
branched <- extract(mv) 
dim(branched)
```

It provides various options to select a subset of data only.
`universe` and `nuni` are useful if you want to select 
a subset of universes in the multiverse. `frow` provides
the option to extract a fraction of data in each universe.

```{r extract_fraction}
frac <- extract(mv, frow =  0.05)
nrow(frac)/nrow(branched)
```

Using the output tibble, you can construct plots to compare
their distributions in different universes. For example,
you can overlay density lines on a single plot.

```{r compare_universe, fig.caption="Density lines overlaid on a single plot.", fig.width=5, fig.height=4, warning=FALSE}
branched %>%
  ggplot(mapping = aes(x = skin_tone, color = universe)) +
  geom_density(alpha = 0.2) +
  xlab('Skin Tone') +
  ylab('Density') +
  ggtitle('Density of player skin tones.')
```

Another option is the use `ggplot`'s `facet_grid` function 
to generate multiple plots in a grid.

```{r compare_universe_hist, fig.caption="Histograms on a facet grid.", fig.width=5, fig.height=4, warning=FALSE}
branched %>%
  ggplot(mapping = aes(x = skin_tone, fill = universe)) +
  geom_histogram(position = 'dodge', bins = 10) +
  labs(title = 'Histogram of player skin tones.',
       y = 'Count', x='Skin Tone') + # you can add axis labels and plot title with labs.
  facet_grid(. ~ universe) # generate histograms for each universe.
```

### Exercise.

1.  Use `geom_boxplot` to compare the distributions between 
the five options.
2.  Construct a grid of histograms for all the five options. 
Use `universe ~ .` instead of `. ~ universe` in `facet_grid`. 

## Discussion.

1.  The five options each result in different distributions. Does one option provide a more _objective_ rating than others?
2.  Among the three types of plots - density plot, histogram, and boxplot, which plot do you find the most effective at comparing different distributions? 


## References
