---
title: "Case Study. Are female hurricanes deadlier than male hurricanes?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study. Are female hurricanes deadlier than male hurricanes?}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
references:
  - id: original
    title: "Female hurricanes are deadlier than male hurricanes"
    type: article-journal
    issued:
      year: 2014
      month: 6
      day: 17
    URL: https://doi.org/10.1073/pnas.1402786111
    DOI: 10.1073/pnas.1402786111
    journal: Proceeedings fo the National Academiy of Sciences of the United States of America
    volume: 111
    issue: 24
    page: 8782-8787
    author:
    - given: Kiju
      family: Jung
    - given: Sharon
      family: Shavitt
    - given: Madhu
      family: Viswanathan
    - given: "Joseph M."
      family: Hilbe
  - id: multiverse
    title: "multiverse: R package for creating explorable multiverse analysis"
    type: entry
    URL: https://mucollective.github.io/multiverse/
    accessed:
      year: 2020
    author:
    - given: Abhraneel
      family: Sarma
    - given: Matthew
      family: Kay
  - id: boba
    title: "Boba: Authoring and Visualizing Multiverse Analyses"
    type: entry
    URL: https://arxiv.org/abs/2007.05551
    issued:
      year: 2020
      month: 7
      day: 30
    accessed:
      year: 2020
    author:
    - given: Yang
      family: Liu
    - given: Alex
      family: Kale
    - given: Tim
      family: Althoff
    - given: Jeffrey
      family: Heer
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE
)
```

This case study is adapted from the `multiverse` package case study of the same analysis.

## Initiate a multiverse

```{r load_data, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(purrr)
library(tidybayes)
library(ggplot2)
library(mverse)
data('hurricane')
# read and process data
hurricane <- hurricane %>%
  rename(
    year = Year,
    name = Name,
    dam = NDAM,
    death = alldeaths,
    female = Gender_MF,
    masfem = MasFem,
    category = Category,
    pressure = Minpressure_Updated_2014,
    wind = HighestWindSpeed
  ) %>%
  mutate(
    post = if_else(year > 1979, 1, 0),
    zcat = as.numeric(scale(category)),
    zmin = -scale(pressure),
    zwin = as.numeric(scale(wind)),
    z3 = as.numeric((zmin + zcat + zwin) / 3)
  )
```

```{r define_mverse}
mv <- create_multiverse(hurricane)
```

## Define filter branches

```{r filter_branch}
death_outliers <- filter_branch(
  TRUE,
  name != 'Katrina',
  !(name %in% c('Katrina', 'Audrey'))
)
damage_outliers <- filter_branch(
  TRUE,
  name != 'Sandy',
  !(name %in% c('Sandy', 'Andrew')),
  !(name %in% c('Sandy', 'Andrew', 'Donna'))
)
mv <- mv %>%
  add_filter_branch(death_outliers, damage_outliers)
summary(mv)
```

## Define mutate branches

```{r define_variables}
femininity <- mutate_branch(
  female,
  masfem
)
damage <- mutate_branch(
  dam,
  log(dam)
)
mv <- mv %>%
  add_mutate_branch(femininity, damage)
summary(mv)
```

## Define model branches

> `mverse` methods aren't implemented yet

```{r fit_model}
# wrapper for model branching to be implemented
multiverse::inside(
  mv, {
    fit <- glm(
      branch(
        model,
        'linear' ~ log(death + 1),
        'poisson' ~ death
      ) ~ branch(
        femininity_damage_interaction,
        'yes' ~ femininity * damage,
        'no' ~ femininity + damage
      ) + branch(
        other_predictors,
        'pressure' ~ femininity * zmin,
        'wind' ~ femininity * zwin,
        'category' ~ femininity * zcat,
        'all' ~ femininity * z3,
        'all_no_interaction' ~ z3,
        'none' ~ NULL
      ) + branch(
        covariates,
        '1' ~ NULL,
        '2' ~ year:damage,
        '3' ~ post:damage
      ), family = branch(
        model,
        'linear' ~ gaussian,
        'poisson' ~ poisson
      ), data = data
    )
    # attach fitted values and standard errors to the original data
    pred <- predict(fit, se.fit = TRUE, type = "response")
    disagg_fit <- data %>%
      mutate(
        fitted = pred$fit,
        se.fit = pred$se.fit,
        df = df.residual(fit),
        sigma = sigma(fit),
        se.residual = sqrt(sum(residuals(fit)^2))/ df
      )
    pred2expectation <- function(mu, sigma) {
      branch(model, "linear" ~ exp(mu + sigma^2/2) - 1, "poisson" ~ mu)
    }
    expectation <- disagg_fit %>%
      mutate(expected_deaths = pred2expectation(fitted, sigma)) %>%
      group_by(female) %>%
      summarise(mean_deaths = mean(expected_deaths), .groups = "drop_last") %>%
      compare_levels(mean_deaths, by = female)
  }
)
```

## Execute and Inspect Results

```{r}
# execute the multiverse
mv <- mv %>%
  execute_multiverse()
```

> `mverse` methods aren't implemented yet (will wait until `multiverse::extract_variables` is available in the released version of `multiverse`)

```{r extract_results, fig.width=8, fig.height=3, fig.cap="Estimated coefficient of femininity of a hurricane's name."}
# a wrapper function to be implemented using multiverse::extract_variables
fem_coefs <- multiverse::multiverse_table(mv) %>%
  mutate(res = map(.results, 'expectation')) %>%
  unnest(res) %>%
  rename(universe = .universe)
fem_coefs %>%
  arrange(mean_deaths) %>%
  mutate(index = 1:nrow(.)) %>%
  ggplot(aes(y = mean_deaths, x = index, color = model)) +
  geom_point(alpha = 0.1) +
  xlab('Universe') + ylab('Mean Difference in Expected Deaths') +
  theme_minimal()
```
